---
title: "covid_FOE_green_space_eda_1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(readr)
library(purrr)
library(DBI)
library(lubridate)
```

https://friendsoftheearth.uk/nature/englands-not-so-green-and-pleasant-land-millions-can-only-access-green-space-size-garden

*Ideas: * 

* K-means clustering: https://www.statsandr.com/blog/clustering-analysis-k-means-and-hierarchical-clustering-by-hand-and-in-r/ Does it come up with different groups to the green space deprivation metrics designed by FOE.



Import total Covid-19 cases for local authorities (as of 08/12/2020). This data was processed in `green_space_eda_3.Rmd`.

```{r}
covid_19_green_space_db <- dbConnect(RSQLite::SQLite(), "covid-19_green_space_db.sqlite")

total_cases <- dbGetQuery(covid_19_green_space_db , "SELECT * FROM total_cases") %>% 
  mutate(date = as_date(date))

dbDisconnect(covid_19_green_space_db )

total_cases
```

What I wanted to be able to do was plot the Friends of the Earth Green Space indicies against the incokme index, but that is not possible as green space indicies are only provided at the MSOA scale.

```{r}
green_space <- read_excel("./Data/(FOE) Green Space Consolidated Data - England - Version 2.1.xlsx",
                          sheet = "Local Authorities V2.1")

green_space
```
```{r}
total_cases %>% 
  select(-date) %>% 
  left_join(green_space, by = c("areaCode" = "LA_Code")) %>% 
  
  ggplot(mapping = aes(x = Income_Index, y = Pcnt_MSOAs_Rating_E, colour = total_cases_per_ht)) +
  
  geom_point()
```

So shift to MSOA scale
```{r}

green_space <- read_excel("./Data/(FOE) Green Space Consolidated Data - England - Version 2.1.xlsx",
                          sheet = "MSOAs V2.1")

green_space

```

[Direct link to the dataset](https://api.coronavirus.data.gov.uk/v2/data?areaType=msoa&metric=newCasesBySpecimenDateRollingSum&metric=newCasesBySpecimenDateRollingRate&metric=newCasesBySpecimenDateChange&metric=newCasesBySpecimenDateChangePercentage&metric=newCasesBySpecimenDateDirection&format=csv) provided by cornavirus.data.gov.uk

```{r}
msoa_cases <- read_csv("./Data/msoa_2020-12-17.csv")
msoa_cases
```

```{r, fig.width=10}
msoa_cases %>% 
  left_join(green_space,
            by = c("areaCode" = "MSOA_Code")
            ) %>% 
  
  select(areaCode, areaName, date, newCasesBySpecimenDateRollingRate, Income_Index, Green_Space_Deprivation_Rating) %>% 
  
  group_by(date, Green_Space_Deprivation_Rating) %>% 
  summarise(average_new_case_rate = mean(newCasesBySpecimenDateRollingRate, na.rm = TRUE)) %>% 
  
  ggplot(mapping = aes(x = date, y = Green_Space_Deprivation_Rating, fill = average_new_case_rate)) +
  
  geom_tile() +
  
  scale_fill_viridis_c()


```


https://www.tidymodels.org/learn/statistics/k-means/
https://www.statsandr.com/blog/clustering-analysis-k-means-and-hierarchical-clustering-by-hand-and-in-r/

```{r}
green_space


```
```{r}

library(tidymodels)
library(NbClust)
library(factoextra)
library(GGally)

```


https://policy.friendsoftheearth.uk/sites/files/policy/documents/2020-10/Green_space_gap_full_report_1.pdf

```{r}

green_space %>% 
  select(GSDI_Access, GSDI_AvgArea, GSDI_Garden) %>% 
  map(~ unique(.x))

green_space_indices <- green_space %>% 
  select(GSDI_Access, GSDI_AvgArea, GSDI_Garden)

green_space_indices
```


```{r}
fviz_nbclust(green_space_indices, kmeans, method = "silhouette")
```


```{r}
set.seed(27)
kclust <- kmeans(green_space_indices, centers = 5, nstart = 100)
summary(kclust)
```

```{r}
green_space <- augment(kclust, green_space)
green_space

green_space %>% 
  group_by(.cluster) %>% 
  summarise(mean_gsdi_access = mean(GSDI_Access),
         mean_gsdi_ave_area = mean(GSDI_AvgArea),
         mean_gsdi_garden = mean(GSDI_Garden),
         med_gsdi_access = median(GSDI_Access),
         med_gsdi_ave_area = median(GSDI_AvgArea),
         med_gsdi_garden = median(GSDI_Garden)
         )
```

```{r}

gsdi_plot <- function(x_var, y_var){
  
  p <- ggplot(data = green_space,
            mapping = aes(x = .data[[x_var]], y = .data[[y_var]], colour = `.cluster`)
            )

  p + geom_jitter()
}

gsdi_plot("GSDI_Access", "GSDI_AvgArea")
gsdi_plot("GSDI_Access", "GSDI_Garden")
gsdi_plot("GSDI_AvgArea", "GSDI_Garden")

```

**What if I repeat the analysis above put use the data that underpins the index/scores create by the FOE?**

There are two measures of green space per capita in the FOE dataset. I guess these were from different sources, but they appear broadly equivalent ...

```{r}

p <- ggplot(data = green_space,
            mapping = aes(x = GreenSpace_Area_Per_Capita, y = Unbuffered_GOSpace_Per_Capita)
            )

p + geom_point(alpha = 0.5)

```
```{r}
green_space_variables <- green_space %>% 
  select(GreenSpace_Area_Per_Capita, Garden_Area_Per_Capita, Pcnt_Pop_With_GOSpace_Access)

green_space_variables
```

```{r}
fviz_nbclust(green_space_variables, kmeans, method = "silhouette")
```

Struggled to get meaningful clusters out in initial experiments (only put around 10 minutes in ...)

```{r}
set.seed(27)
kclust <- kmeans(green_space_variables, centers = 4, nstart = 100)
summary(kclust)

green_space <- augment(kclust, green_space)
green_space

green_space %>% 
  group_by(.cluster) %>% 
  summarise(mean_gs_area =  mean(GreenSpace_Area_Per_Capita), 
            mean_gar_area = mean(Garden_Area_Per_Capita), 
            mean_perc_pop_5_min_walk = mean(Pcnt_Pop_With_GOSpace_Access)
         )

```



Couldn't find data for rural-urban classification - https://www.gov.uk/government/statistics/2011-rural-urban-classification email ons??

