---
title: "covid_FOE_green_space_eda_1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(readr)
library(purrr)
library(DBI)
library(lubridate)
```

https://friendsoftheearth.uk/nature/englands-not-so-green-and-pleasant-land-millions-can-only-access-green-space-size-garden

*Ideas: * 

* K-means clustering: https://www.statsandr.com/blog/clustering-analysis-k-means-and-hierarchical-clustering-by-hand-and-in-r/ Does it come up with different groups to the green space deprivation metrics designed by FOE.



Import total Covid-19 cases for local authorities (as of 08/12/2020). This data was processed in `green_space_eda_3.Rmd`.

```{r}
covid_19_green_space_db <- dbConnect(RSQLite::SQLite(), "covid-19_green_space_db.sqlite")

total_cases <- dbGetQuery(covid_19_green_space_db , "SELECT * FROM total_cases") %>% 
  mutate(date = as_date(date))

dbDisconnect(covid_19_green_space_db )

total_cases
```

What I wanted to be able to do was plot the Friends of the Earth Green Space indicies against the incokme index, but that is not possible as green space indicies are only provided at the MSOA scale.

```{r}
green_space <- read_excel("./Data/(FOE) Green Space Consolidated Data - England - Version 2.1.xlsx",
                          sheet = "Local Authorities V2.1")

green_space
```
```{r}
total_cases %>% 
  select(-date) %>% 
  left_join(green_space, by = c("areaCode" = "LA_Code")) %>% 
  
  ggplot(mapping = aes(x = Income_Index, y = Pcnt_MSOAs_Rating_E, colour = total_cases_per_ht)) +
  
  geom_point()
```

So shift to MSOA scale
```{r}

green_space <- read_excel("./Data/(FOE) Green Space Consolidated Data - England - Version 2.1.xlsx",
                          sheet = "MSOAs V2.1")

green_space

```

[Direct link to the dataset](https://api.coronavirus.data.gov.uk/v2/data?areaType=msoa&metric=newCasesBySpecimenDateRollingSum&metric=newCasesBySpecimenDateRollingRate&metric=newCasesBySpecimenDateChange&metric=newCasesBySpecimenDateChangePercentage&metric=newCasesBySpecimenDateDirection&format=csv) provided by cornavirus.data.gov.uk

```{r}
msoa_cases <- read_csv("./Data/msoa_2020-12-17.csv")
msoa_cases
```

```{r, fig.width=10}
msoa_cases %>% 
  left_join(green_space,
            by = c("areaCode" = "MSOA_Code")
            ) %>% 
  
  select(areaCode, areaName, date, newCasesBySpecimenDateRollingRate, Income_Index, Green_Space_Deprivation_Rating) %>% 
  
  group_by(date, Green_Space_Deprivation_Rating) %>% 
  summarise(average_new_case_rate = mean(newCasesBySpecimenDateRollingRate, na.rm = TRUE)) %>% 
  
  ggplot(mapping = aes(x = date, y = Green_Space_Deprivation_Rating, fill = average_new_case_rate)) +
  
  geom_tile() +
  
  scale_fill_viridis_c()


```


https://www.tidymodels.org/learn/statistics/k-means/
https://www.statsandr.com/blog/clustering-analysis-k-means-and-hierarchical-clustering-by-hand-and-in-r/

```{r}
green_space

green_space %>% 
  select(GSDI_Access, GSDI_AvgArea, GSDI_Garden) %>% 
  map(~ unique(.x))
```
```{r}

library(tidymodels)
set.seed(27)

green_space_indices <- green_space %>% 
  select(GSDI_Access, GSDI_AvgArea, GSDI_Garden)

green_space_indices


kclust <- kmeans(green_space_indices, centers = 5, nstart = 100)
summary(kclust)

augment(kclust, green_space_indices)
```


