---
title: "data_analysis_cases"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Focus on Leeds cases
As I'm interested in Leeds in particular,  I created a tibble of Leeds case data. But also I want to compare the data for Leeds to equivalent national data for England. As the number of cases in Leeds is likely to be a very small proportion of the total number of cases, I have also create a rest of England (i.e. England - Leeds) dataset for comparision.

```{r}

leeds_cases <- tidy_lsoa %>%
  filter(startsWith(lsoa11_nm, "Leeds"))

rest_of_england_cases <- tidy_lsoa %>% 
  filter(!startsWith(lsoa11_nm, "Leeds"))

View(leeds_cases)

```

## Initial plots
```{r}
p_leeds <- ggplot(data = cases_per_week_leeds, mapping = aes(
    x = week_commencing , y = weekly_cases)
  ) +
  geom_col(fill = "#6699FF") +
  scale_x_date(date_breaks = "8 weeks", date_labels = "%d %b") +
  labs(x = NULL, y = "Number of cases (per week)", title = "Leeds") +
  coord_x_date(xlim = c("2020-03-02", "2020-08-24")) +
  theme_minimal()

p_national <- ggplot(data = cases_per_week_rest, mapping = aes(
    x = week_commencing , y = weekly_cases)
  ) +
  geom_col(fill = "grey") +
  scale_x_date(date_breaks = "8 weeks", date_labels = "%d %b") +
  labs(x = NULL, y = NULL, title = "England (exc. Leeds)") +
  coord_x_date(xlim = c("2020-03-02", "2020-08-24")) +
  theme_minimal()

multiplot(p_leeds, p_national, cols = 2)
# plot(p_national)
# plot(p_leeds)
```
