---
title: "green space EDA 2"
output: 
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(purrr)
library(corrplot)
library(colorspace)
library(lubridate)

source("r_code/covid_helper_functions.R")

`%nin%` = Negate(`%in%`)

theme_set(theme_light())
```
Objectives:

* personal interest in the topic
+ green gym
+ covid experience

https://blogs.bmj.com/bmj/2020/07/03/covid-19-has-highlighted-the-inadequate-and-unequal-access-to-high-quality-green-spaces/
https://www.weforum.org/agenda/2020/08/parks-green-spaces-mental-health-access-equality/

* Develop 


Scale of analysis: Local Authority District (the finest geographic granularity available for all datasets)
Geographic scope: England (case data at MSOA level available for England only)

## 1. Import and tidy data sources

### 1.1. Access to private green space (gardens)
The ONS (Office for National Statistics) provides [data on access to private green space](https://www.ons.gov.uk/economy/environmentalaccounts/datasets/accesstogardensandpublicgreenspaceingreatbritain) (i.e. access to gardens) for each Local Authority District in Great Britain. Here I am using the most recent April 2020 edition of the data. I quickly, manually edited the ONS excel file to make it easier use the `read_excel` function for data import. Given it is unlikely that the ONS data will be updated during the course of this analysis, it was preferable to go for the quicker manual process than investing time in a re-producible programmatic approach.

```{r, import gs 1}

gardens <- read_excel("./Data/osprivateoutdoorspacereferencetables.xlsx", sheet = "LAD gardens_2")
gardens %>% 
  sample_n_from_df(10)
```

### 1.2. Access to public green space (parks)
The ONS (Office for National Statistics) also provides [data on access to public green space](https://www.ons.gov.uk/economy/environmentalaccounts/datasets/accesstogardensandpublicgreenspaceingreatbritain) (i.e. access to parks and playing fields) for each Local Authority District in Great Britain. In this case no manual editing of the ONS excel file was required. 

```{r, import gs 2}
parks <-  read_excel("./Data/ospublicgreenspacereferencetables.xlsx", sheet = "LAD Parks only")
parks %>% 
  sample_n_from_df(10)
```

### 1.3. Covid cases
[data.gov.uk](https://coronavirus.data.gov.uk/details/download) provides data on the number of Covid-19 cases (as confirmed by positive tests) occurring in each English Local Authority District. 

```{r, import covid 1, message=FALSE}
# store the date the data was provided by cornavirus.data.gov.uk for use later on in a plot
data_date <- "2020-12-08"
covid_19_file_path <-  str_c("./Data/ltla_", data_date, ".csv")

covid_cases <- read_csv(covid_19_file_path)
                  
covid_cases %>% 
  sample_n_from_df(10)

```


## 1.5. Population data
The ONS also provides [data](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/populationestimatesforukenglandandwalesscotlandandnorthernireland) esitmating the populations resident in Local Authority District. This will be helpful later on when  

```{r}
populations <- read_csv("./Data/ons_mid_2019_population_estimates.csv")
populations

```

## 2. Transform datasets for analysis

### 2.1. Focus down on variables of interest

*Private green space: * The ONS provides data on private green space separately for flats and house, additionally totals (including both flats and houses) are provided. In this case just the totals are likely to be of interest, as I am not expecting to distinguish between types of housing stock in the analysis the relationship between access to green space and Covid-19 case numbers.

```{r}
gardens <- gardens %>% 
  filter(`Country name` == "England") %>% 
  select(`LAD code`, `LAD name`, `T_Address count`, 
         `T_Percentage of adresses with private outdoor space`, 
         `T_Average size of private outdoor space (m2)`
         ) %>% 
  rename(address_count = `T_Address count`,
         perc_with_garden = `T_Percentage of adresses with private outdoor space`,
         ave_garden_size = `T_Average size of private outdoor space (m2)`
         )

gardens
```
*Public green space: * the parks dataset is more straightforward (than the garden's dataset), so I just drop some redudant geographic variables.

```{r}
parks <- parks %>% 
  filter(`Country name` == "England") %>% 
  select(-`Country code`, -`Country name`, -`Region code`, -`Region name`)

parks
```


*Covid_19_cases: * 

* There are 27030 NAs and 21372 zeros in the new_case column. If a NA value for new case is different to a zero value? How can this best be managed ni the analysis below? For now I'll assume an NA can be treated as zero, given there is no documentation on coronavirus.data.gov.uk providing details.
* the number of cases (confirmed by a positive test result) in each Local Authority are given on a daily basis.

```{r}
covid_cases <- covid_cases %>%
  filter(str_detect(areaCode, "^E")) %>% 
  select(-areaType, -newCasesBySpecimenDateRollingSum, -newCasesBySpecimenDateRollingRate) %>% 
  rename(new_cases = newCasesBySpecimenDate)

covid_cases

summary(covid_cases)

covid_cases %>% 
  count(new_cases) %>% 
  head(1)
```
*Population data: * 

* I am interested in the total population of each Local Authority District. So, I drop variables from the `populations` dataframe detailing the breakdown of population by age, and retain the variable for total population within a given geographic area.
```{r}
populations <- populations %>% 
  select(Code, Name, `All ages`) %>% 
  rename(population = `All ages`)

populations
```


### 2.2. Join datasets
Having identified the variables of interest across the four datasets imported above (gardens, parks, covid_cases and populations), I merge these together to form a single dataframe. This will hopefully make things a bit easier later on in terms of visualization, and also help keep the data wrangling processes seperate from subsequent visualization processes.

```{r}
# join the four datasets into a single dataframe
covid_gardens_parks_pop <- covid_cases %>% 
  left_join(gardens, by = c("areaCode" = "LAD code")) %>%
  select(-`LAD name`) %>% 
  left_join(parks, by = c("areaCode" = "LAD code")) %>% 
  select(-`LAD name`) %>% 
  left_join(populations, by = c("areaCode" = "Code")) %>% 
  select(-Name)

```

I then quickly check to see if NA values have been introduced through the joins performed above And, it appears they have. I am not  concerned about the NAs for the `new_cases` variable at this stage, as I've considered these above. 

```{r}

# check how many NAs occur in each column
covid_gardens_parks_pop %>%
  summarise(across(.fns = ~ sum(is.na(.))))

```
But it does look worth checking the extent to which data relating to gardens, parks and population is missing following the join. In particular, how many Local Authorities have missing data?

```{r}
id_LAs_with_NAs <- function(df, column_name){
  
  # convert the string passed to the function for use as a column name in a dplyr function
  column_name_sym <- rlang::sym(column_name)
  
  # identify Local Authorities where an NA is present in the specified column
  df %>%
    filter(is.na(!!column_name_sym)) %>%
    distinct(areaName)
}

# select variable names of interest 
vars <- colnames(covid_gardens_parks_pop %>% 
                   select(-new_cases) # as discussed above there are many NAs for variable
                 )
# create a list of Local Authorities where there are NA entries in the parks, gardens and population datasets
LAs_with_NAs <- vars %>%
  map_df(~ id_LAs_with_NAs(covid_gardens_parks_pop, .x)) %>% 
  distinct()

LAs_with_NAs


```
As the missing data issues are restricted to just five Local Authorities, the issue with NAs appears relatively small and unlikely to adversely impact the analysis I plan to move on to below. I do however remove data for these five local authorities from the aggregated dataframe, to prevent the issue of these local authorities appearing in plot including some variables but not in plots including other variables.

```{r}
covid_gardens_parks_pop <- covid_gardens_parks_pop %>%
  filter(areaName %nin% LAs_with_NAs)
```


### 2.3. Create variables for analysis 
Data on Covid-19 cases confirmed by a positive test within each Local Authority District is recorded on a daily basis. I am trying to understand the overall extent to which each local authority has been affected by Covid-19. So, I calculated total case numbers (up to the 8th December) for each Local Authority District. Additionally, I wanted to take into account the variations in population size across different local authorities. So, I calculated the number of cases per 100,000 population (a metric often used in [government reporting](https://coronavirus.data.gov.uk/details/about-data#rate-calculations)). 

```{r}
total_cases <- covid_gardens_parks_pop %>% 
  
  # calculate total number of cases for each Local Authority District
  group_by(areaCode, areaName) %>% 
  mutate(total_cases = sum(new_cases, na.rm = T)) %>% 
  ungroup() %>% 
  
  # remove date related information and associated duplicate rows as no longer required
  select(-date, -new_cases) %>% 
  distinct() %>% 
  
  # create variables normalized for population
  mutate(cases_per_ht = total_cases / population * 1e+5,
         num_parks_per_ht = `Average number of  Parks within one km` / population * 1e+5) %>% 
  # create other variables of potential interest
  mutate(occupancy_rate = population / address_count)
total_cases
```
## 3. Visualising the relationship between access to green space and Covid-19 case numbers
Here I am mostly interested in identifying any associations between the green space related variables (provided by the ONS), and the rates of Covid-19 cases. So, I begin by producing  a set scatter plots, considering the green space related variables as potential explanatory variables (on the x axis), and the number of cases per 100,000 population as the response variable (on the y axis). Where each point plotted represents an individual Local Authority District.

```{r, fig.width=10}
total_cases_tidy <- total_cases %>% 
  select(-address_count, -population, - total_cases, - num_parks_per_ht, - occupancy_rate) %>% 
  pivot_longer(perc_with_garden:`Average combined size of  Parks within 1km  (m2)`,
               names_to = "explanatory_variable")
  
p1 <-  ggplot(data = total_cases_tidy,
             mapping = aes(x = value, y = cases_per_ht)
             )

p1 + geom_jitter(alpha = 0.5) +
    facet_wrap(~ explanatory_variable, scales="free")

```
From inspection of the scatter plots above, the most promising potential explanatory variables are the average garden size and the average distance to the nearest park. So, focus on these two variables in more detail.

```{r}
vars_of_interest <- c("ave_garden_size", "Average distance to nearest Park (m)")

total_cases_focus <- total_cases_tidy %>% 
  # remove green space variables I am no longer interested in
  filter(explanatory_variable %in% vars_of_interest)
  

total_cases_focus
```

Focusing on average garden size and the average distance to the nearest park, I quickly plotted each variable against cases per  100,000 population, and the log (base 10) of cases per 100,000 population. This was because in the original plots above I had noticed perhaps a hint on non-linearity in the association between the green space variables and the number of cases. Reviewing the plots with the log y scale I thought that this transformation of the data did little to bring what associations are present in better focus. And, log scales are typically less intuitive to interpret than linear scales, so I decided to stick with the linear y scales for the remainder of the analysis.

```{r}
p2 <-  ggplot(data = total_cases_focus,
             mapping = aes(x = value, y = cases_per_ht)
             )

p2 + geom_jitter(alpha = 0.5) +
    facet_wrap(~ explanatory_variable, scales="free")


p3 <-  ggplot(data = total_cases_focus,
             mapping = aes(x = value, y = log(cases_per_ht))
             )

p3 + geom_jitter(alpha = 0.5) +
    facet_wrap(~ explanatory_variable, scales="free")

```

### Exploring the role relative wealth plays in the relationship between access to green space and Covid-19 case numbers
At this point it occurred to me that it would be interesting to explore if the affluence of each Local Authority had a mediating role to play in the relationships between the two green space variables and the number of Covid-19 cases. So, I imported so some  data (again from the ONS) estimating the average salaries of people living in each local authority. Within the dataset two measures of the average are provided: (1) the median salary within a given local authority; and, (2) the mean salary within a given local authority. I retained both as I had yet to make a decision on which was a more appropriate measure in the context of this analysis.


```{r}
income <- read_excel("./Data/PROV - Home Geography Table 8.7a   Annual pay - Gross 2020.xls",
                     sheet = "All_edit_for_import")

income <- income %>% 
  
  # mean and median variables imported as characters (perhaps something to using read_excel ...)
  mutate(Median = as.double(Median),
         Mean = as.double(Mean)
         )
```
A bit of data wrangling was required to integrate the income data into the main dataset I am using for plotting below. Also, I need to remove some NA values that were causing problems in the plotting. I didn't look at these NAs created by coercing the variable types from character to numeric in detail. I just filtered out Local Authority with NA values for the Mean salary later on (ahead of plotting). I also created a couple of new variables which are used later on to facet the plots.

```{r}
# taken from the same dataset as the Local Authority mean/median data
# these are the figures for mean/median salary for England as a whole
eng_median_salary <-  26055	
eng_mean_salary <- 32237
```

```{r}

total_cases_income <- total_cases_focus %>%
  
  # add the income data to the Covid-19 case and green space data
  pivot_wider(names_from = "explanatory_variable", values_from = "value") %>%
  rename(ave_dist_to_park = "Average distance to nearest Park (m)") %>% # simplifying naming of variable of interest
  left_join(income %>% 
              select(Code, Median, Mean),
            by = c("areaCode" = "Code")
            ) %>% 
  
  #create variable to allow grouping in plots below
  mutate(above_mean = Mean > eng_mean_salary,
         above_median = Median > eng_median_salary) %>%
  
  # remove NAs in variables to be included in the plots
  filter(!is.na(cases_per_ht)) %>% 
  filter(!is.na(Mean)) %>% 
  filter(!is.na(Median))

total_cases_income
```




I continued developing the plots of the two green space variables of interest (`ave_dist_to_park` and `ave_garden_size`) vs the total number of Covid-19 cases, by applying colour to provide a basic representation of the relative affluence of each Local Authority District. In the two plots below a diverging colour scale is used: 

*The center of the colour scale (shown as white) is the national median salary; 
*Local authorities with median salaries above the national median are shown in red (the higher above the national median salary of £`r as.integer(eng_median_salary)` the darker the red); 
*Local authorities with median salaries below the national median are shown in blue (the lower below the national median salary of £`r as.integer(eng_median_salary)` the darker the blue).

It took quite a lot of experimentation to find the solution to using a specified midpoint from which the colours diverge. For example, it wasn't obvious how to do this using scale_fill_brewer or if it was even possible ...

In terms of the insights generated from the two plots below:

* In both plots we can see that local authorities with more affluent residents (shown in blue) tend to have experienced fewer cases of Covid-19 per 100,000 population.
* In both plots we can see almost all of the local authorities worst affect by Covid-19 (say those with more than a total of 3500 per 100,000 population) have median salaries below the national median of £`r as.integer(eng_median_salary)`.



```{r}

green_space_covid_plotter <- function(df, x_var, sal_ave, facet_var = NULL, facet_labs = NULL){
  
  # set the midpoint of the diverging colour scale to either England's median or mean salary
  # strengthen the starting colour for the lower half of scale 
  # (as the salary range above the mean is wider than the salary range below)
  
  if (sal_ave == "Median") 
    colour_pallette <- scale_fill_continuous_divergingx(palette = 'RdBu', 
                                        mid = eng_median_salary,
                                        c1 = 0, l1 = 0, p1 = 1, p2 = 1)
  else 
    colour_pallette <- scale_fill_continuous_divergingx(palette = 'RdBu', 
                                        mid = eng_mean_salary,
                                        c1 = 0, l1 = 0, p1 = 1, p2 = 1)
  
  
  # provide the option not to facet the plot
  if(is.null(facet_var)) facet = NULL 
  
  else
    facet = facet_wrap(vars(.data[[facet_var]]), labeller = as_labeller(facet_labs))
  
  # create the plot
  ggplot(data = df, 
             mapping = aes(x = .data[[x_var]],
                           y = cases_per_ht, 
                           fill = .data[[sal_ave]])
             ) +
  
  geom_jitter(size = 2, shape = 21, colour = "grey30") +
  
  colour_pallette +
    
  facet

}

p_park <- green_space_covid_plotter(df = subset(total_cases_income,
                                            subset = ave_dist_to_park < 800), # remove an outlier
                                x_var = "ave_dist_to_park",
                                sal_ave = "Median"
                          )

p_garden <- green_space_covid_plotter(df = subset(total_cases_income, 
                                      subset = ave_garden_size < 1000
                                      ), # remove an outlier
                          x_var = "ave_garden_size",
                          sal_ave = "Median"
                          )

p_park
p_garden
```

I wondered if any relationship between the green space variables might be being obscured by plotting affluent and deprived local authorities on the same figure. I wondered if affluent and deprived local authorities might be somewhat distinct sub-groups. So, 

* In the case of `ave_garden_size` we can see evidence of a negative association between average garden size and the total number of Covid-19 cases. So, local authorities with smaller average garden sizes
* In the case of `ave_dist_to_park` 

Mean versus median

```{r}

# define labels for plot facets
facet_labels_mean <- c(`FALSE` = str_c("Local Authorities with mean salaries\n equal to or below £", eng_mean_salary),
                      `TRUE` = str_c("Local Authorities with mean salaries\n above £", eng_mean_salary))

facet_labels_median <- c(`FALSE` = str_c("Local Authorities with median salaries\n equal to or below £", eng_median_salary),
                      `TRUE` = str_c("Local Authorities with median salaries\n above £", eng_median_salary))



p_park_median <- green_space_covid_plotter(df = subset(total_cases_income,
                                            subset = ave_dist_to_park < 800), # remove an outlier
                                x_var = "ave_dist_to_park",
                                sal_ave = "Median",
                                facet_var = "above_median",
                                facet_labs = facet_labels_median
                          )

p_garden_median <- green_space_covid_plotter(df = subset(total_cases_income, 
                                      subset = ave_garden_size < 1000
                                      ), # remove an outlier
                          x_var = "ave_garden_size",
                          sal_ave = "Median",
                          facet_var = "above_median",
                          facet_labs = facet_labels_median
                          )

p_park_mean <- green_space_covid_plotter(df = subset(total_cases_income,
                                            subset = ave_dist_to_park < 800), # remove an outlier
                                x_var = "ave_dist_to_park",
                                sal_ave = "Mean",
                                facet_var = "above_mean",
                                facet_labs = facet_labels_mean
                          )

p_garden_mean <- green_space_covid_plotter(df = subset(total_cases_income, 
                                      subset = ave_garden_size < 1000
                                      ), # remove an outlier
                          x_var = "ave_garden_size",
                          sal_ave = "Mean",
                          facet_var = "above_mean",
                          facet_labs = facet_labels_mean
                          )



p_park_median
p_garden_median
p_park_mean
p_garden_mean
```



```{r}
# create a new variable to allow faceting by quartile
total_cases_income <- total_cases_income %>% 
  mutate(median_salary_quartile = ntile(Median, 4))

# define labels for plot facets
facet_labs <-  c(`1` = "Q1", `2` = "Q2", `3` = "Q3", `4` = "Q4")
  
  
p_park_median_qs <- green_space_covid_plotter(df = subset(total_cases_income,
                                            subset = ave_dist_to_park < 800), # remove an outlier
                                x_var = "ave_dist_to_park",
                                sal_ave = "Median",
                                facet_var = "median_salary_quartile",
                                facet_labs = facet_labs
                          )

p_garden_median_qs <- green_space_covid_plotter(df = subset(total_cases_income, 
                                      subset = ave_garden_size < 1000
                                      ), # remove an outlier
                          x_var = "ave_garden_size",
                          sal_ave = "Median",
                          facet_var = "median_salary_quartile",
                          facet_labs = facet_labs
                          )
                          
p_park_median_qs
p_garden_median_qs
```

Produce final plot

```{r}
data_date <- ymd(data_date)

x_label <- expression("\nAverage garden size in"~m^2)
y_label <- "Total number of confirmed Covid-19 cases\nper 100,000 population\n"
subtitle <- str_c("The mean salary across England as a whole in 2020 is £", eng_mean_salary, ".\n")
title <- "Average garden size and the number of Covid-19 cases\nin English Local Authorities\n"
legend_label <- "Mean salary in\nthe local authority"

p_garden_mean + labs(x = x_label,
         y = y_label,
         fill = legend_label,
         title = title,
         subtitle = subtitle,
         caption = str_c("\nCovid-19 cases number calculated from data issued on ", format(data_date, "%d/%m/%Y"))
         )
```

## A quick note on what I learnt in the process of producing the plots below

In the process of incorporating the income data into this exploratory data analysis. There were a couple of key areas for me to learn about:  

1. how to [write functions with ggplot2](https://fishandwhistle.net/slides/rstudioconf2020/#1) that can produce plots with common features (e.g. in this case data mapped to the y axis) but with the ability to configure certain aspects plots (e.g. in the case the data mapped to the x axis). It was particularly useful to learn about how adding NULL to a ggplot object is help in creating conditions within functions that produce plots.
2. how to use colour scales for continuous variable. More specifically, [how to set midpoints when using diverging colour scales](https://stackoverflow.com/questions/58718527/setting-midpoint-for-continuous-diverging-color-scale-on-a-heatmap).
3. how to iteratively refactor code throughout the process of create a number of similar plots. 


